Circumnuclear Media and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-02-10 ms.tex
more edits 4.3

Radiative cooling can, however, be stabilized if other sources of cooling dominate over it, in particular the term $\propto -q c_{s}^{2} \propto -qT$ in the entropy equation (eq.~[\ref{eq:dsdt}]).  Neglecting radiative cooling, this term is balanced at the stagnation radius by the wind heating term, $\dot{q}_{\rm heat} = q \tilde{v}_{w}^{2}/2$.  Thermal stability near the stagnation radius can thus be assessed by the ratio of wind heating to radiative cooling $|\dot{q}_{\rm cool}|$ (eq.~[\ref{eq:qdot_rad}]),
\begin{align}
\left.\frac{\dot{q}_{\rm heat}}{|\dot{q}_{\rm rad}|}\right|_{\rm r_{\rm s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhors}) and (\ref{eq:rhostarrs}) for the gas and stellar densities at the stagnation radius, respectively.  We have approximated the cooling function for $T < 2\times 10^{7}$ K as $\Lambda(T) = 1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}$ assuming solar metallicity gas (\citealt{Draine:2011a}; his Fig. 34.1).  To within a constant of order unity, equation (\ref{eq:cooling2}) also equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n kT/2)/\dot{q}_{\rm rad}$ to the free-fall time $t_{\rm ff}$ if one uses the equality 
\begin{align}
\rho|_{r_{\rm s}}=\frac{3 q\tff|_{r_{\rm s}}}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations (\ref{eq:Tanalytic}), \eqref{eq:mdot_analytic}, and~\eqref{eq:mdot_gas}.

Thermal stability ($\dot{q}_{\rm heat} > |\dot{q}_{\rm rad}|$; $t_{\rm cool} > t_{\rm ff}$) at the stagnation radius thus requires a critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}\\
\sigma,     
  \end{cases}
  \label{eq:cooling3}
\end{align}
The stability requirement $v_{\rm w} \gtrsim \sigma$ results because the top two lines of equation (\ref{eq:cooling3}) are derived using equation (\ref{eq:rs_simple}) for the stagnation radius.  However, when $v_{\rm w} \lesssim \sigma$ the stagnation radius moves to much larger radii or may not exist at all, resulting in much higher gas densities and inevitable cooling instability. 
%Our numerical results in $\S\ref{sec:numerical}$ demonstrate that $\tilde{v}_{w} \gtrsim v_{\rm CI} \sim 200-300$ km s$^{-1}$ is generally sufficient for stability at all radii (Fig.~\ref{fig:cooling}).

Substituting condition (\ref{eq:cooling2}) into our analytic expression (\ref{eq:eddr_analytic}) for $\dot{M}(v_{\rm w})$ shows a thermally stable flow possesses a maximum accretion rate, which is given by
\begin{align}
\frac{\dot{M}_{\rm max,TI}}{\dot{M}_{\rm edd}} \simeq
  {\rm min}\begin{cases}
   4\times 10^{-4} \eta_{0.02}^{0.47} M_{\bullet,8}^{0.34} &, \text{core}\\
   2\times 10^{-4} \eta_{0.02}^{0.59} M_{\bullet,8}^{0.29}    &, \text{cusp}\\
\eta M_{\bullet}/\dot{M}_{\rm edd}t_{\rm h} \approx 5.5\times 10^{-5}\eta_{0.02}&, v_{\rm TI} < \sigma.
  \end{cases}
  \label{eq:Mdotmax}
\end{align}

@@

Fig.~\ref{fig:cooling} shows the ratio of wind heating to radiative cooling, $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ (eq.~[\ref{eq:cooling2}] and surrounding discussion) as a function of radius for our fiducial solutions shown in Figure \ref{fig:profiles}.  For high heating rates of $v_{w} = 600$ and $1200$ km s$^{-1}$ cooling is unimportant across all radii, while for $v_{w} = 300$ km s$^{-1}$ we see that $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| < 1$ and $\tcool/\tff \lesssim 1,$ depending on the wind mass loss parameter $\eta$.  Note that the ratio $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ at the stagnation radius deviates from its minimum value across the entire grid by less than a factor of two; thus the ratio of $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}}$ indeed represents a useful proxy for the presence of thermal instabilty across the entire grid.  Also note that for $v_{w}=600$ and $1200$ km s$^{-1}$ we have that $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| \sim t_{\rm cool}/t_{\rm ff}$ near the stagnation radius.  However, the two ratios diverge from each other at low heating ($v_{w} \lesssim 300$ km s$^{-1}$) because equations (\ref{eq:mdot_gas}), (\ref{eq:rhors}) underestimate the true gas density, which is greater in a subsonic inflow (weak heating) than in a supersonic outflow (strong heating).

>>>>

Radiative cooling can, however, be stabilized if other sources of cooling dominate over it, in particular the term $\propto -q c_{s}^{2} \propto -qT$ in the entropy equation (eq.~[\ref{eq:dsdt}]).  Neglecting radiative cooling, this term is balanced at the stagnation radius by the wind heating term, $\dot{q}_{\rm heat} = q \tilde{v}_{w}^{2}/2$.  Thermal stability near the stagnation radius can thus be assessed by the ratio of wind heating to radiative cooling $|\dot{q}_{\rm cool}|$ (eq.~[\ref{eq:qdot_rad}]),
\begin{align}
\left.\frac{\dot{q}_{\rm heat}}{|\dot{q}_{\rm rad}|}\right|_{\rm r_{\rm s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhors}) and (\ref{eq:rhostarrs}) for the gas and stellar densities at the stagnation radius, respectively.  We have approximated the cooling function for $T < 2\times 10^{7}$ K as $\Lambda(T) = 1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}$ assuming solar metallicity gas (\citealt{Draine:2011a}; his Fig. 34.1).  To within a constant of order unity, equation (\ref{eq:cooling2}) also equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n kT/2)/\dot{q}_{\rm rad}$ to the free-fall time $t_{\rm ff}$ at the stagnation radius, as can be derived using the equality 
\begin{align}
\rho|_{r_{\rm s}}=\frac{3 q\tff|_{r_{\rm s}}}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations (\ref{eq:Tanalytic}), \eqref{eq:mdot_analytic}, and~\eqref{eq:mdot_gas}.

Thermal stability ($\dot{q}_{\rm heat} > |\dot{q}_{\rm rad}|$; $t_{\rm cool} > t_{\rm ff}$) at the stagnation radius thus requires a critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}\\
\sigma,     
  \end{cases}
  \label{eq:cooling3}
\end{align}
The requirement $v_{\rm w} \gtrsim \sigma$ results because the top two lines are derived using equation (\ref{eq:rs_simple}) for $r_s$.  However, when $v_{\rm w} \lesssim \sigma$ the stagnation radius moves to very large radii, or may not exist at all, resulting in much higher gas densities and inevitable cooling instability. 
%Our numerical results in $\S\ref{sec:numerical}$ demonstrate that $\tilde{v}_{w} \gtrsim v_{\rm CI} \sim 200-300$ km s$^{-1}$ is generally sufficient for stability at all radii (Fig.~\ref{fig:cooling}).

The existence of a minimum heating rate for thermal stability (eq.~[\ref{eq:cooling2}]) implies a corresponding maximum accretion rate,
\begin{align}
\frac{\dot{M}_{\rm max,TI}}{\dot{M}_{\rm edd}} \simeq
  {\rm min}\begin{cases}
   4\times 10^{-4} \eta_{0.02}^{0.47} M_{\bullet,8}^{0.34} &, \text{core}\\
   2\times 10^{-4} \eta_{0.02}^{0.59} M_{\bullet,8}^{0.29}    &, \text{cusp}\\
\eta M_{\bullet}/\dot{M}_{\rm edd}t_{\rm h} \approx 5.5\times 10^{-5}\eta_{0.02}&, v_{\rm TI} < \sigma,
  \end{cases}
  \label{eq:Mdotmax}
\end{align}
which is derived using equation (\ref{eq:eddr_analytic}) to calculate $\dot{M}(v_{\rm TI})$.

@@

Fig.~\ref{fig:cooling} shows the ratio of wind heating to radiative cooling, $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ (eq.~[\ref{eq:cooling2}] and surrounding discussion) as a function of radius for our fiducial solutions shown in Figure \ref{fig:profiles}.  For high heating rates of $v_{w} = 600$ and $1200$ km s$^{-1}$ cooling is unimportant across all radii, while for $v_{w} = 300$ km s$^{-1}$ we see that $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| < 1$ and $\tcool/\tff \lesssim 1,$ depending on the wind mass loss parameter $\eta$.  At the stagnation radius, the ratio $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ reaches a value that lies within a factor of two of its minimum across the entire grid;  thus, the value of $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}}$ provides a diagnostic of thermal instabilty for the entire solution.

Also note that for $v_{w}=600$ and $1200$ km s$^{-1}$ we have that $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| \sim t_{\rm cool}/t_{\rm ff}$ near the stagnation radius.  However, the two ratios diverge from each other at low heating ($v_{w} \lesssim 300$ km s$^{-1}$) because equations (\ref{eq:mdot_gas}), (\ref{eq:rhors}) underestimate the true gas density, which is greater in a subsonic inflow (weak heating) than in a supersonic outflow (strong heating).
